---
title: "gtaptools"
author: "Thiago Simonato"
date: today
format:
  bookup-html:
    embed-fonts: false
toc: true
---

# Introduction

Welcome to the manual for the *gtaptools* package in R, which is currently under development. The primary objective of this manual is to provide practical examples of using the *gtaptools* package for researchers during the development of research with CGE models.

The package aims to improve file management, increase the produtivity, and allow do build a scriptable pipeline that promote the reproducibility. Also, the package provides tools to graphical visualizations that increase the analytical potential of the database and results.

The *gtaptools* package is designed to be user-friendly and is accompanied by detailed documentation and example code built-in. We hope this manual is useful and welcome feedback and contributions from the community.

```{r}
#| label: setup
#| include: false

library(knitr)
```

# Installation

To use the *gtaptools* package, it's necessary to have R installed on your computer, which can be downloaded from [here](https://www.r-project.org/). Additionally, we recommend downloading RStudio, available at [here](https://posit.co/download/rstudio-desktop/), which provides a user-friendly interface to work with R.

You can install the development version of *gtaptools* from [GitHub](https://github.com/) with:

```{r}
#| label: install
#| echo: true
#| eval: false

# If the devtools package is not already installed, please run the disabled line below.
# install.packages("devtools")
devtools::install_github("tsimonato/gtaptools")

```

## Compatibility

The tools of the package mainly works with files in .har, .sl4 and other formats generated by the [GEMPACK](https://www.copsmodels.com/gempack.htm), which is widely used in Computable General Equilibrium (CGE) modeling[^1]. As a result, some of the package's functionalities require a certain degree of familiarity with the structure of the data used in the GEMPACK suite[^2]. This knowledge is not necessary for users to take full advantage of the package's capabilities in manipulating and visualizing CGE model data, but increase the learning curve. 

[^1]: Many applications can be found at [GTAP portal](https://www.gtap.agecon.purdue.edu/models/research.asp)
[^2]: A detailed technical description of the GEMPACK software can be found in its [anual](https://www.copsmodels.com/gpmanual.htm)

The package also includes functions for data cleaning, manipulation, and visualization that can be used with other data format, like R data.frames. Also, although GEMPACK is only compatible with Windows OS, the package's features work on other OS supported by R and Rstudio such as Linux and MacOS.

# Tools




## Data managment




## File managment



## Data viz




# References













```{r}
#| label: sh
#| echo: true
#| eval: true
say_hello <- function(name) {
    paste("Hello,", name, "!") # <1>
}

say_hello("world") # <2>
```

1.  Quite an original example.
2.  This is were all the action is happening.

### Verbatim

Here is the structure of the `penguins` dataset.

```{r}
#| label: penguins_str
#| echo: true
library(palmerpenguins)
str(penguins)
```

### Tables

Sample table output.

```{r}
#| label: table
tab <- table(penguins$island, penguins$species)
kable(tab)
```

Sample `DT:datatable` output.

```{r}
#| label: dt
library(DT)
DT::datatable(penguins)
```

# Styling

A simple list :

-   one mississipi
-   two mississipi
-   three mississipi

A blockquote :

> Oh ! What a nice blockquote you have here. Much more wonderful than a classical lorem ipsum, really.

And we could also [include links](https://www.r-project.org/) or simply URLs like this : <https://www.r-project.org/>.




An incredibly complex equation :

$$ y = \sqrt{\frac{1}{x + \beta}} $$

# Figures

Here is an histogram.

```{r}
#| label: hist
library(ggplot2)
ggplot(data = penguins) +
    geom_histogram(aes(x = body_mass_g)) +
    facet_grid(rows = vars(species))
```

And a wonderful scatterplot, with a caption.

```{r}
#| label: scatter
#| fig.cap: "This is a scatterplot"
ggplot(data = penguins) +
    geom_point(
        aes(x = bill_length_mm, y = bill_depth_mm)
    )
```

# Callouts

::: callout-note
This is a note callout.
:::

::: callout-warning
This is a warning callout
:::

::: callout-important
This is an important callout
:::

::: callout-caution
This is a caution callout
:::

::: callout-tip
This is a tip callout.With a bit of `code`.
:::

# Tabset

::: panel-tabset
## First tab

This is the first tab content.

## Second tab

This is the second tab content.
:::
